Draft Article

### 2019-12-19 article.tex
Clean up baltic incident

% \section{Aerial espionage}
% \section{Introduction}
% According to \emph{The New York Times}, the very first reconnaissance flight took place in 1911, only eight years after Kitty Hawk, when ``Lieut. Piazzi today, for the first time in the history of warfare, made from this place an aerial reconnaissance against a hostile power'' surveilling the Turkish Infantry near Tripoli.\footcite{special_cable_to_the_new_york_times_air_1911} The ability to fly over an enemy's position to gain intelligence about their movements---in a craft orders of magnitude more versatile than a balloon---was so obviously valuable that countermeasures quickly followed. Early aerial dogfighting famously took the form of pilots firing handguns at each other out of open cockpits in mid-air (emphasis on attempted), but by the time of the First World War, planes were already outfitted with synchronized machine guns and ground regiments equipped with anti-aircraft weaponry. Crucial information about military movements was going to come at a cost.

% Even though these aircraft were performing vital intelligence-gathering functions, that did not make them spy planes. Aerial reconnaissance first appeared in the press in 1911, but the term ``spy plane'' appeared in the pages of \emph{The New York Times} just once before 1960.\footnote{According to my search of the archives. The instance mentioned here was in 1944, when the AP wrote that ``Germans tried to probe the secrets of the gathering Allied storm by E-boat patrol dashes [...] and by spy-plane coastal raids.''(\cite{the_associated_press_britons_1944}) In context of the article, though, it's clear that they are referring to military aerial reconnaissance, especially since it was written during the Second World War. This is a marked difference from a plane created for peacetime intelligence, which is what the term ``spy plane'' has since come to mean.} The image of a spy plane crashed into the public consciousness that year when Gary Powers crawled out of a U-2 class aircraft over Soviet Russia. The downed craft was clearly not a generic fighter plane that had flown too close to the border, and it suddenly became necessary to categorize this new type of plane, flying at previously unheard-of altitudes, that was purpose-built for clandestine photography. The United States first claimed that the lost plane was a ``weather craft.''\footcite{caruthers_soviet_1960} When that cover didn't hold up, a new story was born. The diplomatic fallout of the U-2 shootdown, which threatened to torpedo an upcoming NATO-Warsaw summit, had become the ``spy-plane case.''\footcite[This is the second time that ``spy plane'' as a term of art appeared in \emph{The New York Times}. There would be many more.]{reston_allies_1960}

% I emphasize the linguistic distinction between a military plane used for reconnaissance and a ``spy plane'' because that distinction has a major impact on how the flight is perceived and what diplomatic repercussions might result when it is caught.\footnote{That is not to say there is a hard boundary between military and non-military intelligence, either in purpose or collection. The overlap between military and civilian intelligence agencies---especially in the United States---is considerable. The Air Force was deeply involved with the U-2 Project, for example, even though it was technically under the jurisdiction of the CIA. The aerial photographs taken by U-2 missions were used to inform both military and civilian operations.} A military fight carries with it the implications of combat, aggression, and the possible prelude to an attack. A spy flight is instead a tool of civilian intelligence, whose purposes are often frustrated but never quite discouraged. The creation of the U-2 plane---a craft exclusively for civilian intelligence---is a pivotal moment in the history of espionage because the U-2 is the physical embodiment of the espionage norms it helped to define. Its \emph{raison d'\^etre} is an activity which is not permitted---or its altitude would be unnecessary---but its design emphasizes its peaceful reconnaissance purpose.

% \section{The Open Skies policy}
% From the beginning of the Cold War, aerial intelligence played a crucial role in American foreign policy. The United States considered aerial surveillance so important that it was willing to allow the Soviets to do it in return. President Eisenhower formally proposed a mutual inspection regime called Open Skies to the Soviet Union as a means of reducing security tensions. In his prepared remarks at the Geneva Summit of 1955, Eisenhower laid out the following plan:

% 	% \begin{quote}
% 	\textbf{To give to each other a complete blueprint of our military establishments}, from beginning to end, from one end of our countries to the other, lay out the establishments and provide the blueprints to each other.
% 	\newline

% 	Next, to provide within our countries facilities for aerial photography to the other country---\textbf{we to provide you the facilities within our country, ample facilities for aerial reconnaissance}, where you can make all the pictures you choose and take them to your own country to study, you to provide exactly the same facilities for us and we to make these examinations, and by this step \textbf{to convince the world that we are providing as between ourselves against the possibility of great surprise attack}, thus lessening danger and relaxing tension. (emphasis mine)\footcite{eisenhower_president_1955}
% 	% \end{quote}

% The UN-sponsored Geneva Summit, formally the \emph{Conference on the Peaceful Uses of Atomic Energy}, was a comparatively hopeful period in the Cold War, a conference promoting scientific diplomacy where such a cooperative information exchange might have seemed feasible.\footcite[p.~27]{luscher_nuclear_2018} Nonetheless, the scale of this offer was astounding. Eisenhower offered to let the the Soviets install operational military bases in the United States from which they could surveil American national defense infrastructure, so long as they allowed the US to do the same.\footnote{Also included were reciprocal nuclear facility inspections, an immediate nonstarter in the USSR, but also controversial within the American administration back home. (\cite{prados_review_2015})} In effect, proposed a limited form of state-sanctioned espionage for explicitly defensive purposes---exactly the norm I am arguing exists in practice, albeit in a different form.

% The offer was quickly rejected. In conversation over buffet that evening, Khrushchev politely countered that the plan would not further disarmament at all because it would merely confirm the fragmentary information that their respective intelligence services already possessed.\footcite{department_of_state_memorandum_1955} The response was telling---though it was in no way an endorsement of the activity, it both acknowledged that espionage was taking place between the two superpowers, and that it served a potentially de-escalatory purpose.

% Khrushchev also claimed that the offer was primarily a propaganda move on the part of the US, but backed down when Eisenhower essentially dared him to call his bluff.\footcite[p.~2. This is a secondary source---the book cites a conversation by Eisenhower, quoted in NSC meeting memos that are stored in the Eisenhower library.]{bury_eisenhower_2014} The formal rejection came a few weeks later: ``While giving credit to this proposal's attempt to find a solution,'' USSR Council of Ministers chair Nikolai Bulganin wrote in his report to the Supreme Soviet, ``\textelp{} aerial photography cannot yield the expected results, since both our countries contain limitless expanses in which, if one desires, everything can be hidden.''\footcite{bulganin_meetings_1955} Though a lukewarm excuse, it was not a disdainful one. Earlier in the report, he had even said that the conference, as a whole, ``must be considered a definite success of the peace-loving forces.''

% Was the offer genuine, or was it the stunt that Khrushchev accused it of being? In a book about the relationship between aerial reconnaissance and Eisenhower's New Look foreign policy, historian Helen Bury argues that a willingness to negotiate acceptable and enforceable agreements with foreign powers was central to his Cold War strategy.\footcite[p.~69]{bury_eisenhower_2014} More importantly, the intelligence that an Open Skies policy promised would have been invaluable in his ongoing effort to control defense spending, where Eisenhower was embattled by deeply paranoid American institutions.\footcite[p.~212]{bury_eisenhower_2014} His commitment to Open Skies in practice, however, is still a point of some historical debate. In reviewing her argument, intelligence historian John Prados sees a contradiction between her portrayal of Eisenhower's genuine desire for arms control and Eisenhower's role in significant nuclear buildup.\footcite[p.~233-234]{prados_review_2015} In the following chapter, I will present my own evidence that these two existed as part of the same strategy. Eisenhower used photo-reconnaissance---of the kind that an Open Skies policy would have legally permitted---to create an accurate assessment the military capabilities required to sufficiently deter a Soviet nuclear strike; as a result, the United States built far fewer missiles than it otherwise would have.

% % Valentino: Actually, they were explicitly consistent. Ike thought that nukes were the cheapest way to counter the Soviet numerical advantage in conventional weapons (which were and are much more expensive). So Ike saw both as cost saving measures.

% Regardless of Eisenhower's intent, it is clear that the United States would have far more to gain from such a policy than the USSR, and it made strategic sense for the Soviets to reject the offer. Writing about the history of CIA's imaging program, former imagery analyst David Lindgren notes that Khrushchev had the advantage of huge amounts of publicly-available information about US military installations through newspapers, maps, and official aerial photographs, while Soviet restrictions on press freedom denied symmetrical data.\footcite[p.~38]{lindgren_trust_2000} Allowing for an Open Skies policy would also have revealed that Soviet air defenses were not nearly as far along as their leadership claimed. Khrushchev's son would later write that his father's overriding concern at the time was to conceal Soviet weakness.\footcite[p.~133]{brugioni_eyes_2010}

% Treaty or not, the United States was desperate for intelligence about the Soviet missile program, and aerial photo-reconnaissance was crucial to gaining it. The West's HUMINT capabilities within the Soviet Union were hamstrung by the intense surveillance under which their diplomats were placed, as well as the enormous size of the Russia itself.\footcite[p.~23]{lindgren_trust_2000} As the US increasingly relied on signals intelligence (SIGINT) to gather intelligence, the USSR began pursuing a more aggressive aerial defense policy, demonstrating their willingness to open fire on foreign reconnaissance aircraft. The increased cost associated with these missions did little to deter senior US policymakers from attempting to surveil the skies.\footcite[p.~4]{pedlow_central_1992}

% % The Soviet Union was able to successfully destroy US Military aircraft during purported peacetime because in many cases those US aircraft were not supposed to be there in the first place. In many cases their flight path was a clear violation of Soviet territorial boundaries---a fact of which the administration was aware.\footcite{goodpaster_memorandum_1956}

% In the rest of this chapter, I will argue that Eisenhower satisfied the need for intelligence that an Open Skies treaty would have provided by normalizing the collection of aerial intelligence and minimizing its diplomatic consequences. This process happened in two parts: first, through the routine resolution of military reconnaissance shootdowns, and second, by symbolically divorcing peacetime aerial espionage from military reconnaissance with the creation of a spy plane. For both the military and civilian missions, I will demonstrate how the even the most sensationalized incidents resulted in no more than ineffectual diplomatic protests, even though the spy plane could surveil locations that a ferret flight never could.

% \section{Military reconnaissance flights}
% \subsection{Timeline of early Cold War SIGINT missions}
% Well before the U-2 Incident in 1960, the United States had been routinely flying military aircraft near Soviet airspace, and the USSR had been routinely shooting them down. For a conflict defined by its lack of direct skirmishes between the two superpowers, a surprising number of Americans died in these incidents, which are mostly forgotten to history. A Smithsonian investigation found that over the course of the Cold War more than 200 American pilots were lost to Soviet shootdowns---126 of which remain unaccounted for to this day.\footcite{glenshaw_secret_2017}

% The purpose of these missions was to gather SIGINT identifying the location of critical radar installations along the Soviet border. Converted bombers---called ``ferrets'' by the Air Force---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} According to the CIA, these missions began as early as 1947.\footcite[p.~4]{peterson_maybe_1993} Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were ``overflights,'' missions that deliberately violated Soviet airspace to collect intelligence at a much higher risk of failure. The earliest postwar overflight of the Soviet Union---of which I can find definitive record---took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}

% Determining the quantity of early Cold War reconnaissance flights is difficult because they were a closely guarded secret at the time. Some flights, however, were inadvertently declassified---when they were shot down by enemy air defenses. Since my goal is to understand the diplomatic repercussions of these flights, an effective place to start is a list of reconnaissance flights that attracted diplomatic attention after being shot down. The original shootdown stories are available in the newspaper archives, and the incidents have been helpfully compiled by US Air Force Academy historian Dr. John Farquhar, who specializes in aerial reconnaissance and wrote his master's thesis on how it shaped early Cold War diplomacy. My choice of incidents to analyze in this chapter is based in part on his work.
% % \newline

% Additionally, the CIA history magazine \emph{Cryptologic Quarterly} analyzed the Soviet shootdowns of US reconnaissance aircraft in depth, creating a comprehensive (as far as we know) list of each American overflight that was shot down by the USSR (Table \ref{soviet-shootdowns}). The article was written in 1993, but not declassified until 2009. The author, Michael Peterson, uses two limiting criteria that perfectly suit our purposes: the only flights listed are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) explicitly US reconnaissance aircraft.\footcite[p.~4]{peterson_maybe_1993} Note that not all of these flights are necessarily overflights; many took place over international waters when they were shot down. In effect, the table catalogs the 13 times in which the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5. In the original document, this table lists the first incident as having taken place over the Barents Sea, not the Baltic Sea. Because the description of the mission---including a map of its route in the same document---takes place entirely over the Baltic sea, I have concluded that this has to be a typographical error, and corrected it here.]{peterson_maybe_1993}

% \begin{table}[ht]
% \centering
% \begin{tabular}{llr}
% \textbf{Date}     & \textbf{\makecell[l]{U.S. Service \&\\ Aircraft Type}}   & \textbf{General Location} \\
% 8 April 1950      & USN PB4Y2 Privateer           & Baltic Sea                           \\
% 6 November 1951   & USN P2V Neptune               & Sea of Japan                         \\
% 13 June 1952      & USAF RB-29                    & Sea of Japan                         \\
% 7 October 1952    & USAF RB-29                    & East of Hokkaido/Kuril Is.           \\
% 29 July 1953      & USAF RB-50                    & Sea of Japan                         \\
% 4 September 1954  & USN P2V Neptune               & Sea of Japan                         \\
% 7 November 1954   & USAD RB-29                    & East of Hokkaido.Kuril Is.           \\
% 18 April 1955     & USAF RB-47                    & Off Kamchatka Peninsula              \\
% 10 September 1956 & USAF RB-50                    & Sea of Japan                         \\
% 2 September 1958  & USADF C-130                   & Soviet Armenia (near Turkish border) \\
% 1 May 1960        & CIA U-2                       & Sverdolsk, USSR                      \\
% 1 July 1960       & USAF RB-47                    & Barents Sea                          \\
% 10 March 1964     & USAF RB-66                    & East Germany
% \end{tabular}
% \caption{Summary of Soviet Shootdowns, 1950-1964}
% \label{soviet-shootdowns}
% \end{table}

% I cannot confirm whether Peterson's record is comprehensive. Regardless of how thorough declassified CIA information appears, I cannot rule out the possibility that there were other, more secret missions which for whatever reason the United States government has chosen to continue protecting to this day. Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article are promising, and suggest that the information is relatively complete. First, the article was written for a classified audience as an overview of reconnaissance shootdowns during the 1950s, so it is intended to be comprehensive. Shootdowns are a sufficiently significant event such that it is unlikely the author would have just missed one. Additionally, when the article was declassified 16 years later, it appears to have been declassified without redactions or missing pages. And as with any analysis of diplomatic repercussions and espionage, we are aided by a simple logical trap---if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.

% In the following section, I cross-reference the shootdowns that the CIA says took place with Farquhar's analysis of their media and diplomatic impact. A shootdown is either present in both lists, or just one of the two---some incidents in the CIA list are missing from Farquhar's list because they did not have a diplomatic impact, and some incidents that Farquhar analyzes are not mentioned by Peterson because they were not SIGINT reconnaissance missions and likely (though not definitively) served little espionage function. Together, the full spectrum of aerial reconnaissance incidents demonstrates a clear pattern in which the shootdowns that involved espionage were quickly resolved by both parties, while the shootdown of a flight unrelated to espionage produced swift, significant backlash.

% \subsection{The Baltic Incident}
% The first of these shootdowns, on April 1950, brought the previously-hidden aerial reconnaissance program to the attention of the American people. An unarmed Navy Patrol plane, a PB4Y2 Privateer, was intercepted and downed by Soviet fighters over the Baltic sea. The Soviet ambassador to the US lodged a formal note of protest, incorrectly alleging that the US aircraft which had been shot down was a B-29 bomber.\footcite{kirk_ambassador_1950} Upon receiving the note, Secretary of State Dean Acheson remarked that Ambassador Vishinsky's manner was ``serious but not aggressive or antagonistic,'' and the Secretary ``recommended that publicity on our side should be avoided or, if unavoidable, minimized.''\footcite{kirk_ambassador_1950}

% A few days later, the American ambassador issued a formal response, in which he correctly identified the model of plane and claimed that at no point did it cross into any territorial waters. He blamed the Soviet Union directly, outlining the punishment that the Americans planned to impose:

% % \begin{quote}
% 	The United States Government demands that the the Soviet Government institute a prompt and thorough investigation of this matter in order that the facts set forth above my be confirmed to its satisfaction. The United States Government further demands that the most strict and categorical instructions be issued to the Soviet air force that there be no repetition, under whatever pretext, of incidents of this kind which are so clearly calculated to magnify the difficulties of maintaining peaceful and correct international relationships.

% 	The United States Government fully expects that, when its investigation is completed, the Soviet Government will express its regret for the unlawful and provocative behavior of its aviators, will see to it that those responsible for this action are promptly and severely punished and will, in accordance with established custom among peace-loving nations, pay appropriate indemnity for the unprovoked destruction of American lives and property.\footcite{the_associated_press_text_1950}
% % \end{quote}
% This was a weak response---it imposed no consequences, threatened no sanctions, and demanded only the most \emph{pro forma} of reparations. In essence, the note demanded an apology and a promise not to repeat. There was absolutely no chance that the parties responsible on the Soviet side would be held accountable---in the most serious of cases, some lower-level officers might take the fall, but even that was unlikely to happen here. If the United States had followed through on its demand for reparations---the only material repercussion mentioned---then one could have argued that this statement had some teeth. Instead, the demands were rejected in totality by the USSR.\footcite{salisbury_kremlin_1950} As far as I can tell, no reparations were ever made and no further consequences were imposed as a result.

% Both sides had reasons to make this a larger political issue. The Soviet Union had fired first on a US Navy aircraft---an aircraft which could not have been the aggressor because its only armament was a .45-cal pistol carried by one of the crewmen.\footcite[p.~7]{peterson_maybe_1993} Given that the plane was never found, all available evidence suggests that the flight likely did not violate the traditional 12-mile airspace boundary. Based on the last received position report, today the CIA claims that the plane was flying 20-25 miles off the coast of Latvia at the time it was shot down.\footcite[p.~7]{peterson_maybe_1993} At the time, some unnamed Navy and Air Force officers, speaking to reporters in response to the Soviet protest note, conceded that it was possible that a complete breakdown in the plane's navigation system might have caused it to veer off-course into Latvia, but given its last known position, it would have taken ``a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.''\footcite[Technically, the  article credits the first statement about the possible electronic breakdown to ``Navy and Air Force officials'' and the 90 degrees statement to ``Aerial navigators.'' It is unclear whether the navigators in questions are with the US military, but it does not matter regardless.]{the_new_york_times_soviet_1950}

% % \begin{figure}
% % \centering
% % \includegraphics[scale=0.45]{baltic-flight-path.png}
% % \caption{USN PB4Y2 Flight Path}
% % \label{baltic-flight-path}
% % \end{figure}

% Meanwhile, the United States knew that it hadn't lost a B-29, and that the missing plane probably hadn't ended up in Soviet airspace, but the Privateer in question \emph{had} been performing a questionably legal reconnaissance mission when it was shot down. Though the US was likely correct that the plane never flew over Latvia, it certainly got as close as it needed to in order to gather the necessary intelligence about Soviet military installations on the coast. No traditional boundaries were violated, but a repurposed Navy privateer loaded with electronic reconnaissance equipment flying close to Latvia looked, to use a less technical term, shady. Yet based on the initial exchange of diplomatic protests, neither side was inclined to magnify the issue if they could avoid it.

@@

% The most important consequence of the Baltic Incident was not diplomatic. As a direct result of the shootdown, President Truman, who until that point was largely uninformed about these missions, ordered a thirty-day halt to all reconnaissance flights. In that time, his office drafted an operating procedure that would balance the strategic need for reconnaissance with its potential diplomatic volatility.\footcite[p.~41]{farquhar_aerial_2015} The resulting memo from the Joint Chiefs of Staff created, with Truman's approval, the Special Electronic Airborne Search Project (SESP), which set clear delineations of responsibility between Navy and Air Force reconnaissance, and established executive responsibility for approving the missions.

@@

% \subsection{Routine resolution}
% Between the shootdown of the Navy Privateer in 1950 and the famous U-2 shootdown in 1960, the United States lost nine more aircraft to Soviet air defenses. Without analyzing each one at the same level of detail, I can demonstrate how the general attitudes displayed in response to the Baltic Incident became normalized. In each of the incidents listed, the pattern is the same---each side released dueling statements of protest, and the situation continued more or less as before.

% Describe here are the next few peacetime reconnaissance shootdowns whose diplomatic consequences Farquhar analyzes. A B-29 that went missing on October 7, 1952 hit the front page of the \emph{New York Times} two days later. This mission is on Peterson's list as a ferret flight. Farquhar notes that the incident happened during the US election season and that the media thought the attack was intended to lower American prestige, but he does not mention any diplomatic consequences.\footcite[p.~43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website. As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions ``of which little is known''.\footcite[p.~45]{farquhar_aerial_2015}

>>>>

\section{Aerial espionage}
The line between a \enquote{cold} and a \enquote{hot} war is supposed to be obvious. A superpower's own uniformed military personnel opening fire on the other superpower's military or territory would be \enquote{hot,} everything else is \enquote{cold.} Broadly speaking, history remembers the Cold War as lacking direct military confrontation between the United States and Russia; though the two superpowers of the Cold War were open rivals, they supposedly competed for influence by funding, arming, or otherwise influencing proxies, rather than attacking each other directly. That impression is not exactly accurate. Not only did the US and USSR often come close to openly attacking each other, but a surprising number of times they actually did—--including one of the Cold War's most famous moments.

% citation needed on the definition of cold

Pilot Francis Gary Powers knew that U-2 overflights of the Soviet Union were incredibly risky. The Soviets were deeply unhappy that the US was violating their airspace in such a brazen fashion, but for years they had kept quiet about the U-2 missions, embarrassed that they were technologically incapable of intercepting the high-flying aircraft.\footcite[p.~59]{powers_operation_2004} When Soviet air defenses finally managed to shoot one down in 1960, Soviet leadership was equal parts triumphant and furious. At a planned summit in Paris only a few weeks later, Khrushchev launched into a rant about the U-2 planes and stormed out on the first day, effectively ending the summit---and any hope that pressing issues like the fate of West Berlin would be resolved there. He also canceled a planned visit by Eisenhower to the USSR that was scheduled to take place the next month. A contemporary newsreel about the conference said that \enquote{in the course of two hours, Khrushchev brought US-Soviet relations to their lowest point since the end of World War II.}\footcite{universal_studios_summit_1960}

The shootdown of Gary Powers, what would later come to be known as \enquote{The U-2 Incident}, clearly violated the only \enquote{rule} of the Cold War. Official Soviet air defenses shot down a US aircraft. All parties involved had cause to escalate military---the Soviets for the violation of their airspace, the US for the attack on their \enquote{weather craft}---but instead they chose to treat the U-2 missions as an intelligence operation, not military action. The Soviets therefore applied the traditional consequences associated with espionage; Khrushchev engaged in some saber-rattling and denounced the violation of his airspace, but no material actions were taken, and the United States escaped the incident with few tangible consequences. The striking disparity between the diplomatic furor and the lack of material consequences can be explained by an apparent technicality: the U-2 was not a military aircraft, it was a spy plane.

No single attribute defines a spy plane. A military aircraft can forgo weapons but, by virtue of its pilot, its mission, its timing, or its branding still remain in its operation unquestionably military. Inversely, intelligence missions can be flown by uniformed Air Force pilots in re-purposed B-52s and somehow still be read as espionage. A spy plane is an ontological designation, one that reveals itself to the adversary through a series of social and physical signifiers that are not replicable across time. Germany in 1937 would not have reacted to a U-2 overflight the same way that the Soviet Union did in 1960. But even if the way that different states handled espionage was perfectly consistent, there still wouldn't be a checklist that guarantees a mission will treated as espionage; it is not enough make a military mission look like an intelligence operation. The reality is both simpler and more frustrating---it has \emph{to be one}.

The U-2 incident was the most visible in a series of aerial intelligence operations that the United States undertook in the early Cold War. The US did not always employ civilian pilots nor it did always use obviously innocuous aircraft, yet it managed to run a shocking number of missions that violated Soviet airspace without significantly altering the relative level of tension between the two superpowers. On occasion, these missions would turn \enquote{hot} when Soviet air defenses shot them down. Not once did these shootdowns lead to retaliation from the US, or significant pressure from the Soviets to stop the flights. Instead, each one was resolved as if the pilots who survived were simply spies, captured in the normal course of counterintelligence. This article seeks to answers how and why states ensure that certain operations reveal themselves as espionage; the decade of covert aerial espionage was one such successful operation.

\subsection{Military reconnaissance flights}
The tense battle between the US and USSR over surveillance of Soviet border installations was never formally acknowledged by either party---and neither were its combatants. Over the course of the Cold War, more than 200 American pilots were shot down while spying on the Soviet Union.\footcite{glenshaw_secret_2017} These missions were so highly classified that in many cases the families of the pilots never learned how they died. 126 of these American airmen remain unaccounted for today, casualties of missions whose existence both the US government and the Soviet Union collaborated for decades to deny.

Though they flew Air Force planes, these men were actually conducting one of the earliest forms of Cold War peacetime espionage. Their purpose was to gather SIGINT (signals intelligence) identifying the location of critical radar installations along the Soviet border. Converted bombers---called \enquote{ferrets} by the Air Force---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} According to the CIA, these missions began as early as 1947.\footcite[p.~4]{peterson_maybe_1993} Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were \enquote{overflights,} missions that deliberately violated Soviet airspace. The earliest postwar overflight of the Soviet Union---of which I can find definitive record---took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}

Without access to classified archives it is impossible to state for certain exactly how many of these flights were conducted. Even if one did have unlimited access, some of the necessary information was so highly classified that it is likely lost forever. Nonetheless, the civilian researcher is aided by the declassified research of military archivists before them. In this paper, I highlight the work of two scholars in particular. First, Air Force Academy historian Dr. John Farquhar specializes in aerial reconnaissance and its diplomatic consequences. Many of the incidents analyzed in this section are taken from newspaper clippings that he compiled and analyzed.

Second, former intelligence analyst and section chief Michael Peterson wrote an article for the CIA history magazine \emph{Cryptologic Quarterly} that analyzed the Soviet shootdowns of US reconnaissance aircraft in depth. The article was written for a classified audience in 1993 and declassified in 2009, providing the public with a comprehensive (as far as we know) list of all American SIGINT flights shot down by the USSR (Table \ref{soviet-shootdowns}). Peterson uses two limiting criteria that perfectly suit the purposes of this article: the only flights he lists are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) explicitly US reconnaissance aircraft.\footcite[p.~4]{peterson_maybe_1993} Note that not all of these flights are necessarily overflights; many took place over international waters when they were shot down. In effect, the table catalogs the 13 times in which the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5. In the original document, this table lists the first incident as having taken place over the Barents Sea, not the Baltic Sea. Because the description of the mission---including a map of its route in the same document---takes place entirely over the Baltic sea, I have concluded that this has to be a typographical error, and corrected it here.]{peterson_maybe_1993}

\begin{table}[ht]
\centering
\begin{tabular}{llr}
\textbf{Date}     & \textbf{\makecell[l]{U.S. Service \&\\ Aircraft Type}}   & \textbf{General Location} \\
8 April 1950      & USN PB4Y2 Privateer           & Baltic Sea                           \\
6 November 1951   & USN P2V Neptune               & Sea of Japan                         \\
13 June 1952      & USAF RB-29                    & Sea of Japan                         \\
7 October 1952    & USAF RB-29                    & East of Hokkaido/Kuril Is.           \\
29 July 1953      & USAF RB-50                    & Sea of Japan                         \\
4 September 1954  & USN P2V Neptune               & Sea of Japan                         \\
7 November 1954   & USAD RB-29                    & East of Hokkaido.Kuril Is.           \\
18 April 1955     & USAF RB-47                    & Off Kamchatka Peninsula              \\
10 September 1956 & USAF RB-50                    & Sea of Japan                         \\
2 September 1958  & USADF C-130                   & Soviet Armenia (near Turkish border) \\
1 May 1960        & CIA U-2                       & Sverdolsk, USSR                      \\
1 July 1960       & USAF RB-47                    & Barents Sea                          \\
10 March 1964     & USAF RB-66                    & East Germany
\end{tabular}
\caption{Summary of Soviet Shootdowns, 1950-1964}
\label{soviet-shootdowns}
\end{table}

Regardless of how thorough declassified CIA information appears, I cannot rule out the possibility that there were other, even more secret missions which for whatever reason the United States government has chosen to continue protecting to this day. Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article are promising, and suggest that the information is relatively complete. First, the article was written for a classified audience as an overview of reconnaissance shootdowns during the 1950s, so it is likely intended to be comprehensive. Shootdowns are a sufficiently significant event such that it is unlikely the author would have just missed one. Additionally, when the article was declassified 16 years later, it appears to have been declassified without redactions or missing pages. And as with any analysis of diplomatic repercussions and espionage, we are aided by a simple logical trap---if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.

To comprehensively analyze the diplomatic effects of aerial espionage, I cross-referenced the shootdowns that the CIA says took place with Farquhar's analysis of their media and diplomatic impact. A shootdown is either present in both lists, or just one of the two---some incidents in the CIA list are missing from Farquhar's list because they did not have a diplomatic impact, and some incidents that Farquhar analyzes are not mentioned by Peterson because they were not SIGINT reconnaissance missions and likely (though not definitively) served little espionage function. Together, the full spectrum of aerial reconnaissance incidents demonstrates a clear pattern in which shootdowns that were interpreted as espionage were quickly resolved by both parties, while the shootdown of a flight unrelated to espionage produced swift, significant backlash.

The first of these shootdowns, on April 1950, brought the previously-hidden aerial reconnaissance program to the attention of the American people. An unarmed Navy Patrol plane, a PB4Y2 Privateer, was intercepted and downed by Soviet fighters over the Baltic sea. The Soviet ambassador to the US lodged a formal note of protest, incorrectly alleging that the US aircraft which had been shot down was a B-29 bomber.\footcite{kirk_ambassador_1950} A few days later, the American ambassador issued a formal response, in which he correctly identified the model of plane and claimed that at no point did it cross into any territorial waters. He demanded from the Soviets \enquote{a prompt and thorough investigation} and that the Soviet Government \enquote{see to it that those responsible for this action are promptly and severely punished.}\footcite{the_associated_press_text_1950} The Soviet Union recognized this for the toothless statement that it was, and rejected the totality of its demands.\footcite{salisbury_kremlin_1950} No further action was taken.

Both sides had reasons to make this a larger political issue. The Soviet Union had fired first on a US Navy aircraft---an aircraft which could not have been the aggressor because its only armament was a .45-cal pistol carried by one of the crewmen.\footcite[p.~7]{peterson_maybe_1993} Given that the plane was never found, all available evidence suggests that the flight likely did not violate the traditional 12-mile airspace boundary. Based on the last received position report, today the CIA claims that the plane was flying 20-25 miles off the coast of Latvia at the time it was shot down.\footcite[p.~7]{peterson_maybe_1993} At the time, it was theorized that a theoretical breakdown in the plane's navigation system could have caused it to veer off-course into Latvia, but given its last known position, it would have taken ``a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.''\footcite{the_new_york_times_soviet_1950} Meanwhile, the United States knew that the missing plane probably hadn't ended up in Soviet airspace, but the Privateer in question \emph{had} been performing a questionably legal reconnaissance mission when it was shot down. No traditional boundaries were violated, but a repurposed Navy privateer loaded with electronic reconnaissance equipment flying close to Latvia looked, to use a less technical term, shady. And still, based on the initial exchange of diplomatic protests, neither side was inclined to magnify the issue if they could avoid it.

@@

@@

The general attitudes displayed in response to the Baltic Incident became the normal course of response to Soviet flights. The shootdown of the Navy Privateer was not an isolated incident; from then until the U-2 Incident in 1960, the United States lost nine aircraft to Soviet air defenses. Each time, the pattern of responses was the same---each side released dueling statements of protest, and the situation continued more or less as before. A B-29 that went missing on October 7, 1952 hit the front page of the \emph{New York Times} two days later. This mission is on Peterson's list as a ferret flight. Farquhar notes that the incident happened during the US election season and that the media thought the attack was intended to lower American prestige, but he does not mention any diplomatic consequences.\footcite[p.~43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website. As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions ``of which little is known''.\footcite[p.~45]{farquhar_aerial_2015}

In another incident on July 29, 1953, the United States protested the shootdown of a B-50 (reconnaissance flight) over the Sea of Japan. The Soviet Union responded by protesting the alleged American shootdown of a Soviet passenger plane. Despite press outrage, no further action was taken.\footcite[p.~47]{farquhar_aerial_2015} A September 4, 1954 shootdown of a Navy P2V Neptune actually led to a call by one US Senator to break off diplomatic ties with the Soviet Union: ``Just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers \textelp{} that this new attack upon an American plane confirms Communist arrogance and aggressiveness to a point where the breaking of diplomatic relations is justified.''\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets. The move demonstrated Eisenhower's ability to resist the calls for harsher action while still placating the American public.\footcite[p.~47]{farquhar_aerial_2015}
